Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-07-21 article.tex
modif from DE

The human brain is a complex biological
organ made of about 100 billions of neurons, each connected to 7000 other neurons on
average \citep{pakkenberg2003aging}. Unfortunately, direct observation of the connectome, the wiring
diagram of the brain, is not yet technically feasible. Without being perfect,
calcium imaging currently allows the real-time and simultaneous observation of
neuron activities from thousands of neurons, producing individual time series
representing their fluorescence intensity. From these data, the connectome
inference problem amounts to retrieving the synaptic connections between
neurons on the basis of the fluorescence time series. This
problem is often difficult because of experimental issues, including
masking effects (i.e., some of the neurons are not observed or confounded with others), the low sampling rate of the optical device with respect to the
neural activity speed or the slow decay of fluorescence.

@@

As Figure~\ref{fig:original_curve} shows, the raw
fluorescence signal is very noisy due to light scattering artifacts that
usually affect the quality of the recording~\citep{lichtman2011big}.
Accordingly, the first step of our pipeline is to smoothen the signal, using
one of the following low-pass filters for filtering out high frequency noise:
\begin{align}
% symmetrical median filter
f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i, \label{eq:symetric-median} \\
% asymmetrical weighted median
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
\label{eq:weighted-asymetric-median}
\end{align}
These filters are standard in the signal processing field \citep{kaiser1977data, oppenheim1983signals}. The idea of the second one is to not consider future values and to introduce decreasing weights. For the sake of illustration, the effect of the filter $f_1$ on the signal
is shown in Figure \ref{fig:lp_curve}.

Furthermore, short spikes, characterized by a high
frequency, can be seen as an indirect indicator of neuron communications while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i, \label{eq:high-pass-filter}
\end{align}
as shown in Figure \ref{fig:hp_curve}.

@@

Obvioulsy, Figure~\ref{fig:pca} shows that for both approaches only 800
components are required to explain the (majority of) network structures.
Therefore, the value of the parameter $M$ is immediat and should be clearly
set to $800$, or equivalently to $0.8p$. \\

\begin{figure}[t]
\centering
\subfigure[PCA]{\includegraphics[width=0.45\textwidth]{explained_variance_all} \label{fig:PCA}}
\subfigure[SVD]{\includegraphics[width=0.45\textwidth]{singular_values_all} \label{fig:SVD}}
\caption{Explained variance ratio by component (left) and singular value ratio by component (right) for all networks.}
\label{fig:pca}
\end{figure}

>>>>

The human brain is a complex biological
organ made of about 100 billion of neurons, each connected to 7000 other neurons on
average \citep{pakkenberg2003aging}. Unfortunately, direct observation of the connectome, the wiring
diagram of the brain, is not yet technically feasible. Without being perfect,
calcium imaging currently allows the real-time and simultaneous observation of
neuron activities from thousands of neurons, producing individual time series
representing their fluorescence intensity. From these data, the connectome
inference problem amounts to retrieving the synaptic connections between
neurons on the basis of the fluorescence time series. This
problem is often difficult because of experimental issues, including
masking effects (i.e., some of the neurons are not observed or confounded with others), the low sampling rate of the optical device with respect to the
neural activity speed or the slow decay of fluorescence.

@@

As Figure~\ref{fig:original_curve} shows, the raw
fluorescence signal is very noisy due to light scattering artifacts that
usually affect the quality of the recording~\citep{lichtman2011big}.
Accordingly, the first step of our pipeline is to smoothen the signal, using
one of the following low-pass filters for filtering out high frequency noise:
\begin{align}
% symmetrical median filter
f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i, \label{eq:symetric-median} \\
% asymmetrical weighted median
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
\label{eq:weighted-asymetric-median}
\end{align}
These filters are standard in the signal processing field \citep{kaiser1977data, oppenheim1983signals}. For the sake of illustration, the effect of the filter $f_1$ on the signal
is shown in Figure \ref{fig:lp_curve}.

Furthermore, short spikes, characterized by a high
frequency, can be seen as an indirect indicator of neuron communication while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i, \label{eq:high-pass-filter}
\end{align}
as shown in Figure \ref{fig:hp_curve}.

@@

Obvioulsy, Figure~\ref{fig:pca} shows that for both approaches only 800
components are required to explain the (majority of) network structures.
There, we can conclude that the value of $M$ should be set to 800, or equivalently, to $0.8 p$.

Therefore, the value of the parameter $M$ is immediat and should be clearly
set to $800$, or equivalently to $0.8p$. \\

\begin{figure}[t]
\centering
\subfigure[PCA]{\includegraphics[width=0.45\textwidth]{explained_variance_all} \label{fig:PCA}}
\subfigure[SVD]{\includegraphics[width=0.45\textwidth]{singular_values_all} \label{fig:SVD}}
\caption{Explained variance ratio by number of principal components (left) and singular value ratio by number of principal components (right) for all networks.}
\label{fig:pca}
\end{figure}
